---
title: "R Notebook"
output: html_notebook
---

Methods in Biostatistics with R 
Ciprian Crainiceanu, Brian Caffo, John Muschelli

#1. Introduction

"....We provide a modern look at introductory biostatistical concepts and associated computational tools, reflecting the latest developments in computation and vi- sualization using the R language environment (R Core Team 2016). The idea is to offer a complete, online, live book that evolves with the newest developments and is continuously enriched by additional concepts..."

1.1 Biostatistics 

"....Biostatistics is the theory and methodology for the acquisition and use of quantitative evidence in biomedical research. 
Biostatisticians develop innovative designs and analytic methods targeted at increasing available information, 
improving the relevance and validity of statistical analyses, making best use of available information and communicating relevant uncertainties....”

"....Biostatistics is a challenging subject to master if one is interested in understanding both the underlying concepts and their appropriate implementation. 
....Biostatistics can be viewed as a powerful and practical philosophy of science, in which the scientific hypothesis, 
the experiment, the data, the model, and the associated inference form the basis of scientific progress. 
....Biostatistics is hard because it is focused on solving difficult problems using simple approaches and simplicity is difficult to achieve in complex scenarios.
....Biostatistics is not hard in the usual sense that it requires either deep mathematical concepts or committing large books to memory...." 

"....The difficulty of biostatistics rests in the different philosophy, the different way of thinking, 
and the different set of skills required to process human communication of scientific problems and their translation into well-defined problems that can be solved with data. 
Once a scientific promlem is defined, the biostatistics philosophy is to try to solve it using the simplest possible approaches that are not too simplistic. 
Thus, parsimony and inductive reasoning are fundamental concepts in biostatistics...."

"....biostatistics starts with the current accepted state of knowledge (the collection of null hypotheses) and uses data to inductively refute or reinforce
parts of the knowledge or generate new potential knowledge (new null hypotheses). In the biostatistical philosophy there is no truth, just the state of current knowledge...."

"....Biostatistics requires: 
(1) a tight coupling of the biostatistical methods with ethical and scientific goals of research; 
(2) an emphasis on the scientific interpretation of statistical evidence to impact policy; and 
(3) a detailed acknowledgment of assumptions and a comprehensive evaluation of the robustness of conclusions to these assumptions...."

1.4 Data 
All data located for this course can be downloaded on the leanpub book page(https://leanpub.com/biostatmethods) or can be downloaded directly from GitHub(https://github.com/muschellij2/biostatmethods).





#Chapter 2 Introduction to R


2.4 Objects In R

"....Almost everything is an “object” or “variable.”....most things are relevant only with respect to a data set...."

2.5 Assignment
"....There is also a “forward assignment” operator that exists, but this is rarely used and we do not recommend using it...."
Oh no.

2.6 Data types In R

"....The three most fundamental: numeric (numbers), character (strings or “words”), and logicals (TRUE or FALSE). 

"....One additional data type: factors. Factors are categorical data types. 
The categories of a factor are called the “levels” of that factor. The levels of factors usually contain human-readable levels (e.g., “Male” and “Female”). 
Factors become increasingly useful when we fit models, as one can define the baseline category by the levels of the factor, which affects the interpretation of coefficients. 
Some refer to vectors of length 1 as a “scalar”. In other languages, there is a distinction between scalars and vectors, but not in R. 
So you can think of a vector as a 1-dimensional object that can have anywhere from 0 to a large number of elements...."

2.7.1 More than one number: vectors In R

"....Anything having one dimension is generally referred to as a “vector”. A vector has a specified length. Technically, a vector may have length 0, which means nothing is in it. 
But usually vectors have one or multiple objects in them. Those objects all must have the same data type (e.g., all numbers or all characters or all logicals)...."

2.7.3 Sequences

````{r}
1:5
-5:1
-(5:1)
seq(1,5)
seq(1,7 by = 0.4)
seq(by = 0.4, to = 7, from = 1)


````


2.7.4 Operations on numeric vectors

````{r}
x_length_2 = c(1, 2) 
x_length_2 * 3
sqrt(x_length_2)

````


2.7.5 Some standard operations on vectors 

````{r}
x_length_3 = c(10, 20, 30)
length(x_length_3)
x = c(2, 1, 1, 4)
sort(unique(x))
sample(x)
sample(x, size = 2)

# sample(x, size = 10) will error 
sample(x, size = 10, replace = TRUE)

````



2.7.7 Data type coercion 
"....Vectors can only have one data type in them...."

"....Logicals can be coerced to numerics (0 and 1 values) and coerced to characters ("TRUE" and "FALSE", note the quotes). 
Numerics can be coerced into characters (5.6 changes to "5.6")...."

You can determine the data type of an object with the typeof() function: 

````{r}
typeof(c(5.2, TRUE, FALSE))

````

2.7.8 More than one dimension: Matrices 

"....A matrix is like a vector, but it has two dimensions, rows and columns. Like vectors, matrices can contain only one data type...."
So I imagine a vector is just 'A row' and a matrix is 'A row and a column', or something like that

````{r}
matrix(1)
matrix(1:6, nrow = 2, ncol = 3) 
matrix(1:6, nrow = 2, ncol = 3, byrow = TRUE) # filled by row

````

2.7.9 More than one dimension: data.frames 

"....A data.frame is like a matrix in that it has rows and columns. A data.frame can have columns of different data types...."

....Note, a default on the data.frame() function is stringsAsFactors = TRUE, which indicates that all strings should be converted to factors. 
Many times, you want to clean these columns and then convert them to factors directly. 
So we will set that option to FALSE...."


````{r}
df = data.frame(age = c(25, 30, 32, 42), 
                handed = c("left", "right", "ambidextrous", "left"), 
                stringsAsFactors = FALSE)
````


2.7.10 Dimensions of an object 

There are functions that get attributes about a matrix/data.frame that are useful for analysis. 

The dim function returns the dimensions (rows and columns, respectively) of a data set: 
````{r}
df = data.frame(age = c(25, 30, 32, 42), 
                handed = c("left", "right", "ambidextrous", "left"), 
                stringsAsFactors = FALSE)
dim(df) 

````

Number of rows or columns
````{r}
df = data.frame(age = c(25, 30, 32, 42), 
                handed = c("left", "right", "ambidextrous", "left"), 
                stringsAsFactors = FALSE)
nrow(df)
ncol(df)

````

2.8 Logical operations 

"....Perform some logical tests that return logical data. 
greater/less than (>, <)
greater/less than or equal to (>=, <=)
equal to (== “double equals”)
and not equal to (!=)
These are called “relational operators” or “comparison operators”

There are also the standard “logical operators”: 
not/negate (!), 
AND (&), and 
OR (|)...." 

````{r}
x = c(1, 3, 4, 5) 
x < 4
x > 3
x >= 3
x == 3
x != 3
x = c(1, 3, 4, 5) 
x > 3 | x == 1 
!(x > 3 | x == 1) # negates the statement - turns TRUE to FALSE [1] FALSE TRUE FALSE FALSE 
x != 3 & x != 5
x == 3 | x == 4

````

2.8.1 The %in% Operator 

"....Many times, we want to keep or exclude based on a number of values...." 

````{r}
x = c(1, 3, 4, 5) 
x %in% c(3,4)

````
This feels like an "is in?" or may even an "is?" in that it asks: "are any of these values in the vector?".



2.9 Subsetting 2.9.1 Subsetting vectors

````{r}
x = c(1, 3, 4, 5)
x[4]
x > 2
x[ x > 2 ]

````

2.9.2 Subsetting matrices

"....the syntax is [row subset, column subset]...."

````{r}
mat = matrix(1:6, nrow = 2)
mat[1:2, 1]
mat[1, 1:2]
mat[1, c(FALSE, TRUE, FALSE)]
mat[, c(FALSE, 3, TRUE)]
mat[1:2, ]
mat[, 2:3]
mat[, c(FALSE, TRUE, FALSE)]

````

"....You can ensure that the result is a matrix, even if the result is one-dimensional using the drop = FALSE argument...."
````{r}
mat[,1, drop = FALSE]

````


2.12.1 Tibbles 

Implemented in the tibble package. It is like a data.frame
• Never coerces inputs (i.e. strings stay as strings!). 
• Never adds row.names. 
• Never changes column names. 
• Only recycles length 1 inputs. (no wraparound) 
• Automatically adds column names.

````{r}
tbl = tibble::as_tibble(df)

````



2.13 Problems 



Problem 1. Consider the vector: num <- c("1.5", "2", "8.4")
a. Convert num into a numeric vector using as.numeric 
````{r}
num <- c("1.5", "2", "8.4")
as.numeric(num)

````

b. Convert num into a factor using factor, calling it num_fac. 
````{r}
num <- c("1.5", "2", "8.4")
num_fac <- factor(num)

````

c. Convert num_fac into a numeric using as.numeric. 
````{r}
as.numeric(num_fac)

````

d. Convert num_fac into a numeric using as.numeric(as.character()) 
````{r}
as.numeric(as.character(num_fac))

````



Problem 2. Consider the vector: num <- c(0, 1, 1, 0) 
a. Append TRUE to the num vector using c. 
````{r}
num <- c(0, 1, 1, 0) 
num <- c(num, TRUE)
num

````

b. Check the class of num using class. 
````{r}
num <- c(0, 1, 1, 0) 
class(num)

````

c. Convert num into a logical vector using as.logical. 
````{r}
num <- c(0, 1, 1, 0) 
as.logical(num)
````

d. Convert num into a logical vector where the result is TRUE if num is 0, using the == operator. 
````{r}
num <- c(0, 1, 1, 0) 
as.logical(num == 0)

````



Problem 3. Consider the data set data_bmi from above. Perform the following operations: 
a. Extract the column nams of data_bmi using the colnames() function. 
````{r}
file_name = file.path("Example_Data/bmi_age.txt")
data_bmi = read.table(file = file_name, header = TRUE, stringsAsFactors = FALSE) # found elsewhere in the book. Downloaded from the repository
colnames(data_bmi)

````

b. Subset the AGE column using data_bmi[, "AGE"]. 
````{r}
file_name = file.path("Example_Data/bmi_age.txt")
data_bmi = read.table(file = file_name, header = TRUE, stringsAsFactors = FALSE)
data_bmi[, "AGE"]

````

c. Subset the AGE column using data_bmi$AGE. 
````{r}
file_name = file.path("Example_Data/bmi_age.txt")
data_bmi = read.table(file = file_name, header = TRUE, stringsAsFactors = FALSE)
data_bmi$AGE

````

d. Subset the AGE and BMI columns using brackets and the c() function.
````{r}
file_name = file.path("Example_Data/bmi_age.txt")
data_bmi = read.table(file = file_name, header = TRUE, stringsAsFactors = FALSE)
data_bmi[c("AGE", "BMI")]

````




Problem 4. Here we will work with sequences and lengths: 
a. Create a sequence from 1 to 4.5 by 0.24. Call this run_num. 
````{r}
run_num <-seq(from = 1, to = 4.5, by = 0.24)


````

b. What is the length of run_num. 
````{r}
length(run_num)

````

c. Extract the fifth element of run_num using brackets.
````{r}
run_num[5]

````


Problem 5. Let’s create a tibble called df:
a. Extract the column x using the $.
````{r}
df = dplyr::data_frame(x = rnorm(10), y = rnorm(10), z = rnorm(10)) 
df$x

````

b. Extract the column x using the [,] notation.
````{r}
df = dplyr::data_frame(x = rnorm(10), y = rnorm(10), z = rnorm(10)) 
df[,"x"]






````

c. Extract columns x and z. 
````{r}
df = dplyr::data_frame(x = rnorm(10), y = rnorm(10), z = rnorm(10)) 
df[c("x", "y")]

````

d. Extract the third and fifth rows of df and columns z and y.
````{r}
df = dplyr::data_frame(x = rnorm(10), y = rnorm(10), z = rnorm(10)) 
df[c(3,5),]
df[c("x","y")]
````




Problem 6
a. Get the mean of the column x using the $ operator. 

````{r}
df = dplyr::data_frame(x = rnorm(10), y = rnorm(10), z = rnorm(10))
mean(df$x)



````


b. Pipe (%>%) the column x into the mean function. 

````{r}
library(dplyr)
df$x %>% mean

````

c. Look at the summarize function to look ahead to future chapters and try to do this in the dplyr framework.

````{r}
print(paste("K"))
````

Problem 7.
Consider the data set data_bmi using the BMI data, but read in using readr: 

````{r}
file_name = "Example_Data/bmi_age.txt"
data_bmi = readr::read_table2(file = file_name) 
````
a. What is the class of data_bmi? 
b. What is the class of data_bmi[, "AGE"]? 
c. What is the class of data_bmi$AGE? 
d. What is the class of data_bmi[, "AGE", drop = TRUE]?

